A Match Made in Landfills?

### 2023-04-08 5 Manuscript/manuscript/Manuscript.qmd
BibTEX added

---
title: "A Match Made in Landfills?"
subtitle: "Exploring the diversity and burden of antimicrobial resistance genes carried by white stork (Ciconia ciconia) throughout the breeding season in Madrid, Spain"
author: Raquel Francisco and Seth Latter
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

*We have 126 observations taken over a period of 2 years. We are evaluating for the presence and burden of antimicrobial resistance genes in white stork feces. Codebook is a WIP located in raw data folder.*

@@

*To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:*

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

@@

@@

*Study Area* --- This study took place in Prado Herrero, a private cattle ranch located northwest of metropolitan Madrid and is surrounded by agriculture (e.g. beef cattle, cereal grains, legumes, and forage plants ("Renta Agraria," 2019)). Prado Herrero is located within a nationally protected area (Cuenca Alta del Manzanares Regional Park) and is just north of Santillana Reservoir. This reservoir that was declared an Important Bird Area by Regional Catalogue of Reservoirs and Wetlands of the Community of Madrid due to the numerous resident and migratory species that utilize this water source <CITE>. This cattle ranch has supported a productive white stork rookery where storks have been banded and monitored by biologists at UCM for over 20 years \<CITE: Aguirre and Atienza\>. During the 2020 to 2021 breeding season, between the months of March to June, stork nests were identified, marked, and monitored for productivity. Of marked nests between the 2020 and 2021 breeding seasons, 31 with banded adults were used both years to lay eggs successfully. All 31 nests were located in ash trees (Fraxinus angustifolius ) found within the cattle pasture. Storks that breed within the Prado Herrero rookery are known to utilize Colmenar Viejo Landfill which is located approximately 12km southeast. Colmenar Viejo is an open-air landfill and it is second largest of it's kind in the Madrid region \<CITE: Alejandro 2022\>.

*Sample Collection* --- Between March to May 2020 and 2021, we collected feces from marked nests with banded adults in known breeding pairs at 3 points of the breeding season; (1) an adult sample during incubation, (2) an early juvenile sample during the first two weeks of the chicks life when adults are believed to forage on natural sources \<CITE: \>, and (3) a late juvenile sample after chicks were past two weeks of age when adults forage on anthropogenic resources. Nests were visited in the late mornings and approximately one gram of fresh feces was collected from the perimeter of the nest structure into a sterile Eppendorf tube (). Samples were maintained cold in a portable cooler with frozen gel packs and frozen in a −20°C freezer within 4 hours of collection and processed at a later date.

@@

We performed total DNA extraction directly from fecal samples, by using a pressure filtration technique (QuickGene DNA Tissue Kit S, Fujifilm, Japan) following the manufacturer's instructions. The 16S rRNA gene was amplified in each DNA sample by real time PCR (rtPCR) in 10-fold dilutions of extracted samples, according to Jiang et al. (2013)<CITE:Fix>. A DNA sample was considered validated when a ten-fold dilution showed a cycle threshold (Ct) less than 25 (Esperón et al., 2018). Once validated, we analyzed DNA samples by with a panel of 21 different ARGs encoding resistance to eight different antimicrobial classes: tetracyclines (tet(A), tet(B), tet(Y), tet(K), tet(M), tet(Q), tet(S), and tet(W)), sulfonamides (sulI and sulII), aminoglycosides (str and aadA), phenicols (catI and catII), macrolides (ermB and ermF), quinolones (qnrS and qnrB), betalactams (blaTEM and mecA), and polymyxins (mcr-1). All rtPCR reactions utilized premade gelled format 96-well plates (Biotools, B &M Labs, S.A., Madrid, Spain), with the exception of blaTEM and mecA genes which used the Sybr GreenTM and TaqManTM probe, respectively. Our thermal cycle was the same for all the rtPCR reactions \[6′ 95 °C, 40× (10″ 95 °C, 30″ 60 °C)\], with alignment and extension in the same step, at constant temperature of 60 °C. A melting curve step was performed at the end of the qPCR reaction to validate the authenticity of the positive (Nieto-Claudin et al., 2019). We quantified the relative burden of each gene for each sample via the cycle threshold (Ct) for the 16S rRNA and the Ct value of the individual ARG using a previously published formula (Esperón et al., 2020).

@@

Presence absence ARG results obtained from the fecal samples between 2020 to 2021 were used for simple summary statistics. Samples were classified as "multiresistant" if they were resistant to three or more of the 8 antibiotic classes that we evaluated for in this study (Blanco-Peña et al., 2017). In addition, we applied the following formula to estimate the percentage of bacteria harboring ARGs: x = 10\[2+0.33(ct16S-ctARG)\], where x individual percent gene burden in the sample (i.e., the estimated number of copies of the gene present per reaction). Results were expressed in log10, ranging from −8 ( zero to a negligible amount of the bacteria in the sample carried an ARG) to 2 (all or 100% of the bacteria in the sample carried an ARG). The inverse Log10 was then applied to results so they could be totalled and used to evalute total gene burdens accross sampling periods.

Several linear mixed models (LMM) were constructed to evaluate multi-resistance and ARG burden as response variables with nest as a random factor. Covariates considered with each response variable included adult age, adult mean land fill use index (LUI), sample period (as described above), and nest success. Landfill use was quantified by physically observing a banded stork at Colmenar Viejo during weekly visits from March to June in 2021. The LUI was calculated as the number of observations of one particular bird within the total number of visits to landfill per year \<CITE Alejandro 2022\>. If a banded adult was not seen at the landfill during the breeding season, they were assigned a LUI of 0, suggestive of no resource provisioning at the landfill. All covariates were evaluated for correlation, no covariates were correlated with all the Spearman's correlation coefficients (rho) \< 0.5 and the p \> 0.05. All continuous variables (LUI, age, and nest success) were then standardized prior to analysis.

@@

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("3 Analysis Scripts","LUIvsBurden_LME.tiff"))
knitr::include_graphics(here("3 Analysis Scripts","LUIvsMDR_LME.tiff"))
```

@@

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
ModelAIC1 = readRDS(here("4 Results", "Tables", "ModelAIC1.rds"))
ModelAIC1
knitr::kable(ModelAIC1)

@@

LUI appears to be correlated with higher levels of AMR gene burden in storks. As LUI increases thorough the breeding season (Bialas et al 2021) resistance gene burden also increases with beta lactam resistance contributing to the majority of the burden. However, multidrug resistance appears to decrease as LUI increases, thus it is likely that storks are being exposed to antimicrobial resistance genes at other foraging areas (urban centers, agricultural pastures, etc.). Our results suggest that landfills may not be contributing significantly to the emergence and maintenance of AMR in this system. Little literature exists on the relationship between stork habitat selection and health outside of landfill use in Spain. Future efforts in stork AMR research should focus on exploring the relationship between agricultural land use and health.

@@

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

>>>>

---
title: "A Match Made in Landfills?"
subtitle: "Exploring the diversity and burden of antimicrobial resistance genes carried by white stork (Ciconia ciconia) throughout the breeding season in Madrid, Spain"
author: Raquel Francisco and Seth Latter
date: "`r Sys.Date()`"
format: 
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: EW_AMR.bib
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
library(here)
library(knitr)
library(knitcitations)
```

@@

## General Background Information (WILL FIGHT WITH BIBTEX TOMORROW)

Antimicrobial resistance (AMR) in wildlife was recently highlighted as being a critical research need [@Dolejska2019]. While AMR is accepted as a global human health concern, it has only recently been utilized to evaluated wildlife health and anthropogenic impacts on environments (i.e., environmental health). The majority of resistance research in wildlife has been targeted towards birds, however, the importance of birds in respect to the epidemiology of AMR remains poorly understood [@Radhouani2014]. Now, a growing body of literature is trying to understand the role free-roaming birds play in the emergence, maintenance, and transmission of AMR upon the landscape.

Birds have a long history of being used as indicators for environmental and human health. For example, wading bird surveillance is commonplace in south Florida to monitor heavy metal levels in the greater everglades ecosystem. Birds are also often utilized in zoonotic pathogen surveillance (i.e., USDA sentinel chickens to monitor West Nile Virus). In recent years, a similar approach has been utilized to study AMR prevalence/impacts on different landscapes. Currently, the carriage of AMR genes presents an unknown risk to the individual birds and may have greater implications for free-living avian communities and conservation. Resistance genes are thought to alter the fitness of a pathogen [@Friedman2016], and have been correlated with increased virulence [@Escudeiro2019], in turn increasing morbidity and mortality in affected hosts. Finally, at the population management level, birds have also been held ‘responsible’ for the dissemination of enteric pathogens (e.g., *Escherichia coli*, *Salmonella* spp.) that have caused outbreaks in produce, yet, they are only moving these bacteria from areas already contaminated by human activities [@Hoelzer2011]. Thus, understanding how avian behaviors that overlap with risky environmental factors (e.g., foraging in agricultural areas or landfills) impact the carriage of resistance, is key because: 1) it allows us to better understand the role birds play in AMR dissemination, and  2) it informs targeted population management and conservation strategies—such as discouraging birds from utilizing certain areas.

In 2019, the USGS revealed that more than half of GPS-tracked gulls (*Larus* ssp.) had acquired antimicrobial resistant *E. coli* from landfills and disseminated it long distances to pristine, unaltered habitats [@Ahlstrom2019]. This supports the belief that birds who utilize and move between human-dominated and natural landscapes may become both reservoirs and efficient transporters of resistance determinants to natural environments.For wildlife in general, resistance prevalence tends to increase with proximity to human populations [@Niu2013]. It is believed that animals that rely on anthropogenic resources are often concentrated at high densities that promote the large-scale mixing of bacteria, encouranging horizontal gene transfer (HGT) facilitating the rise of novel resistant strains [@Wellington2013]. For instance, birds that encounter human waste (e.g., by drinking contaminated water) can acquire resistance genes through the exchange of naturally present genes in the bacteria they carry, with the genes in the waste—this can co-select for mobile genetic elements carrying multiple resistant genes [@Wellington2013]. Moreover, ecological factors such as migration and high densities during breeding season may increase acquisition rates and the dissemination of AMR . The western white stork are one such species that heavily utilizes agricultural and urbam areas, providing a unique opportunity to explore how anthropogenic land use influences the carriage of antimicrobial resistance genes.

White storks (*Ciconia ciconia*) were historically trans-Sahara migrants whose ecology has dramatically changed in response to anthropogenic pressure. Since the 1990s, Spanish white stork populations have steadily increased, due to their ability to exploit anthropogenic habitats, especially in areas with abundant resources especially open landfills [@Cramp1985; @Tortosa2002]. Currently, colonies in Madrid, Spain with increased proximity to landfills have improved breeding outcomes and nestling quality, however eggs and nestlings also have increased pollutant loads [@Vergara2006; @Jiang2013; @Saez2008; @Tortosa2002]. This has in turn affected their natural history. Subsets of the population have begun to shorten or abandon their winter migration [@Gordo2007; @Tortosa1995; @Vergara2004]. Although the conservation status of the white stork is of Least Concern in Spain, it has disappeared from large areas of its historical range in Europe and thus, the populations in Portugal and Spain are is integral to the species conservation.

Most studies on AMR surveillance in wildlife are performed by culture‐dependent methods, (e.g., the isolation of a specific pathogen such as *E. coli*) as indicators of AMR. However, since most bacteria are not culturable, the detection of AMR using traditional methods might not be representative of the whole resistant microbiota, sometimes referred to as the resistome [@Surette2017]. Real time PCR (rtPCR), a technique not dependent on culture, offers that ability to quantify the presence and abundance of antimicrobial resistance genes (ARGs), in white storks. This technique has recently been used in three studies to investigate the presence and amount of ARGs in the environment (e.g., soil and manure) and wildlife (e.g., Galapagos tortoise, gulls, guignas) [@Esperon2014; @NietoClaudin2019; @Sacristan2020]. In this study we will explore the ARG load and diversity in a population of white stork in Madrid Spain during the 2020 and 2021.

@@

@@

### Description of data and data source

We have 126 observations taken over a period of 2 years. We are evaluating for the presence and burden of antimicrobial resistance genes in white stork feces. Codebook is a WIP located in raw data folder.

@@

*Study Area* --- This study took place in Prado Herrero, a private cattle ranch located northwest of metropolitan Madrid and is surrounded by agriculture (e.g. beef cattle, cereal grains, legumes, and forage plants. Prado Herrero is located within a nationally protected area (Cuenca Alta del Manzanares Regional Park) and is just north of Santillana Reservoir. This reservoir that was declared an Important Bird Area by Regional Catalogue of Reservoirs and Wetlands of the Community of Madrid due to the numerous resident and migratory species that utilize this water source. This cattle ranch has supported a productive white stork rookery where storks have been banded and monitored by biologists at UCM for over 20 years [@Aguirre2009]. During the 2020 to 2021 breeding season, between the months of March to June, stork nests were identified, marked, and monitored for productivity. Of marked nests between the 2020 and 2021 breeding seasons, 31 with banded adults were used both years to lay eggs successfully. All 31 nests were located in ash trees (*Fraxinus angustifolius*) found within the cattle pasture. Storks that breed within the Prado Herrero rookery are known to utilize Colmenar Viejo Landfill which is located approximately 12km southeast. Colmenar Viejo is an open-air landfill and it is second largest of it's kind in the Madrid region [@Lopez-Garcia2021].

*Sample Collection* --- Between March to May 2020 and 2021, we collected feces from marked nests with banded adults in known breeding pairs at 3 points of the breeding season; (1) an adult sample during incubation, (2) an early juvenile sample during the first two weeks of the chicks life when adults are believed to forage on natural sources, and (3) a late juvenile sample after chicks were past two weeks of age when adults forage on anthropogenic resources. Nests were visited in the late mornings and approximately one gram of fresh feces was collected from the perimeter of the nest structure into a sterile Eppendorf tube. Samples were maintained cold in a portable cooler with frozen gel packs and frozen in a −20°C freezer within 4 hours of collection and processed at a later date.

@@

We performed total DNA extraction directly from fecal samples, by using a pressure filtration technique (QuickGene DNA Tissue Kit S, Fujifilm, Japan) following the manufacturer's instructions. The 16S rRNA gene was amplified in each DNA sample by real time PCR (rtPCR) in 10-fold dilutions of extracted samples, according to @Jiang2013. A DNA sample was considered validated when a ten-fold dilution showed a cycle threshold (Ct) less than 25 [@Esperon2020]. Once validated, we analyzed DNA samples by with a panel of 21 different ARGs encoding resistance to eight different antimicrobial classes: tetracyclines (tet(A), tet(B), tet(Y), tet(K), tet(M), tet(Q), tet(S), and tet(W)), sulfonamides (sulI and sulII), aminoglycosides (str and aadA), phenicols (catI and catII), macrolides (ermB and ermF), quinolones (qnrS and qnrB), betalactams (blaTEM and mecA), and polymyxins (mcr-1). All rtPCR reactions utilized premade gelled format 96-well plates (Biotools, B &M Labs, S.A., Madrid, Spain), with the exception of blaTEM and mecA genes which used the Sybr GreenTM and TaqManTM probe, respectively. Our thermal cycle was the same for all the rtPCR reactions \[6′ 95 °C, 40× (10″ 95 °C, 30″ 60 °C)\], with alignment and extension in the same step, at constant temperature of 60 °C. A melting curve step was performed at the end of the qPCR reaction to validate the authenticity of the positive [@NietoClaudin2019]. We quantified the relative burden of each gene for each sample via the cycle threshold (Ct) for the 16S rRNA and the Ct value of the individual ARG using a previously published formula in @Esperon2020.

@@

Presence absence ARG results obtained from the fecal samples between 2020 to 2021 were used for simple summary statistics. Samples were classified as "multiresistant" if they were resistant to three or more of the 8 antibiotic classes that we evaluated for in this study [@Blanco-Pena2017]. In addition, we applied the following formula to estimate the percentage of bacteria harboring ARGs: x = 10\[2+0.33(ct16S-ctARG)\], where x individual percent gene burden in the sample (i.e., the estimated number of copies of the gene present per reaction). Results were expressed in log10, ranging from −8 ( zero to a negligible amount of the bacteria in the sample carried an ARG) to 2 (all or 100% of the bacteria in the sample carried an ARG). The inverse Log10 was then applied to results so they could be totaled and used to evaluate total gene burdens across sampling periods.

Several linear mixed models (LMM) were constructed to evaluate multi-resistance and ARG burden as response variables with nest as a random factor. Covariates considered with each response variable included adult age, adult mean land fill use index (LUI), sample period (as described above), and nest success. Landfill use was quantified by physically observing a banded stork at Colmenar Viejo during weekly visits from March to June in 2021. The LUI was calculated as the number of observations of one particular bird within the total number of visits to landfill per year [@Lopez-Garcia2021]. If a banded adult was not seen at the landfill during the breeding season, they were assigned a LUI of 0, suggestive of no resource provisioning at the landfill. All covariates were evaluated for correlation, no covariates were correlated with all the Spearman's correlation coefficients (rho) \< 0.5 and the p \> 0.05. All continuous variables (LUI, age, and nest success) were then standardized prior to analysis.

@@

```{r}
#| label: fig-result
#| fig-cap: "Linear Mixed Model Output with 95% CI."
#| echo: FALSE
knitr::include_graphics(here("3 Analysis Scripts","LUIvsBurden_LME.tiff"))
knitr::include_graphics(here("3 Analysis Scripts","LUIvsMDR_LME.tiff"))
```

@@

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Model AICc Tables."
#| echo: FALSE
ModelAIC1 = readRDS(here("4 Results", "Tables", "ModelAIC1.rds"))
ModelAIC1
knitr::kable(ModelAIC1)

@@

LUI appears to be correlated with higher levels of AMR gene burden in storks. As LUI increases thorough the breeding season [@Bialas2020] resistance gene burden also increases with beta lactam resistance contributing to the majority of the burden. However, multidrug resistance appears to decrease as LUI increases, thus it is likely that storks are being exposed to antimicrobial resistance genes at other foraging areas (urban centers, agricultural pastures, etc.). Our results suggest that landfills may not be contributing significantly to the emergence and maintenance of AMR in this system. Little literature exists on the relationship between stork habitat selection and health outside of landfill use in Spain. Future efforts in stork AMR research should focus on exploring the relationship between agricultural land use and health.

@@
